Open-source tools in R for landscape ecology \thanks{Grants or other notes about the article that should go on the front page should be placed here. General acknowledgments should be placed at the end of the article.}

### 2021-01-05 paper/paper.tex
Add linenumbers (Modulo 5)
ljg edits

\institute{
        Maximillian H.K. Hesselbarth \at
     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA \\
     \email{\href{mailto:mhk.hesselbarth@gmail.com}{\nolinkurl{mhk.hesselbarth@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Jakub Nowosad \at
     Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland \\
     \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Johannes M. Signer \at
     Department of Wildlife Sciences, University of Göttingen, Büsgenweg 3, 37077 Göttingen, Germany \\
     \email{\href{mailto:j.m.signer@gmail.com}{\nolinkurl{j.m.signer@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Laura J. Graham \at
     Geography and Environment, University of Southampton, Highfield Rd, Southampton SO17 1BJ, United Kingdom \\
     \email{\href{mailto:l.graham@bham.ac.uk}{\nolinkurl{l.graham@bham.ac.uk}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows using the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
This democratize scientific research as free-of-cost software removes one gatekeeper for researches without access to proprietary software.

@@

The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review, we focus on software implemented in the \emph{R} programming language.
For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
Because in addition to handling spatial data, also tasks such as statistical modeling, creation of publication-ready figures, and even preparation of complete reports can be done in the same working environment (Fig. \ref{fig:fig-workflow}), the \emph{R} programming language can be a very powerful tool.
Thus, in the first part, we present an overview of existing \emph{R} packages for landscape ecology (Table \ref{tab:packages}).
In the second part, we present a survey in which we asked the community how they currently use the \emph{R} programming language and to identify topics for which \emph{R} packages are presently missing for landscape ecology.

@@

In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
Currently, the most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
A \emph{raster} successor, \emph{terra}, aimed at the simpler interface and improved performance is now being developed \cite{R-terra}; however, it could take several years for this package to be adopted by other developers and users.
Alternatively, the \emph{stars} package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
Additionally, there are packages that improve some basics raster operations in terms of computational performance or compatibility between raster and vector operations, such as \emph{fasterize} \cite{Ross2020}, \emph{rasterDT} \cite{OBrien2020}, or \emph{exactextractr} \cite{Baston2020}.

@@

Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
A lot of this data is publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \emph{R} session.
Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows to download data from the Landsat 8, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
Also, the \texttt{getData()} function from the \emph{raster} package allows to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package \emph{geodata} ).
Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps .
This is also referred to as the spatial projection and is often the first barrier in spatial data analysis.
It is not only required to have all of the used data in the same projection, but also to select a proper CRS.
This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion, and different CRS are optimized for different properties, regions of the world, and scales .
Coordinates in spatial data represent one of many coordinate reference systems.
Two main groups of CRS, namely geographical and projected, exist, with each having many members.
In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
However, most landscape ecology studies should utilize projected CRSs, which use some measurement units (e.g., meters).
The selection of projected CRS should be based on the property of spatial data that we want to keep intact (e.g., does not distort areas, shapes, distances, or angles) and be appropriate for a given study area.
Tools to find an appropriate CRS for a certain region can be found at \url{https://spatialreference.org}, \url{https://epsg.org}, or \url{http://epsg.io}.
All packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations, allowing to unify spatial projections when the used data have different CRS.

A second common spatial data processing task is required when the available data extends over a larger area than the study region.
In this case, the pre-processing of spatial data should include vector clipping or raster cropping.
Related to that, masking certain areas of the study region can be required by using, e.g., certain spatial filters such as water bodies.
Packages from Section \ref{sec:spatrep} also allow for these operations.
Additionally, they offer many other operations, such as merging or joining spatial data, extracting values from one dataset into another, raster resolution changes, or vector data simplifications.
A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the \emph{grainchanger} package \cite{Graham2019}.
Furthermore, \emph{landscapetools} is a collection of various utility functions for the raster data model \cite{Sciaini2018}.

Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine. Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.

@@

The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.
Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{R-mapview}.
Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.

@@

One of the most fundamental steps of landscape ecology analyses is to describe and quantify landscape characteristics \cite{Turner2005,Lausch2015}.
For discrete land cover classes, the composition (number and abundance) and configuration (spatial arrangement) of the landscape are often described using landscape metrics \cite{Gustafson1998,Uuemaa2009,Uuemaa2013,Gustafson2019}.
These metrics allow to compare different landscapes, quantify temporal and spatial landscape changes and investigate interactions between landscape characteristics and ecological processes \cite{Uuemaa2009}.

The introduction of the \emph{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019}.
However, while \emph{FRAGSTATS} is free to download at the developers' homepage, it is not open-source software.
This restricts its transparency, reproducibility, and extensibility.
Recently the \emph{landscapemetrics} package \cite{Hesselbarth2019a} was developed to provide a truly open-source solution to calculate landscape metrics.
The package allows calculating the most widely used landscape metrics in a transparent and reproducible workflow within the \emph{R} environment.

More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
The \emph{geodiv} package \cite{Smith2020} allows to calculate gradient surface metrics to characterize continuous analysis of landscape features.
Additionally, the \emph{belg} package allows calculating the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.

Most landscape metrics are a single number depicting specific characteristics of a local landscape.
Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
Spatial signatures calculated for many landscapes can be compared using one of the existing distance measures.
This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
All of the methods mentioned above are implemented in the \emph{motif} package \cite{R-motif}.

@@

Species distribution modeling (SDM) examine how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species distribution, mainly to infer ecological processes and predict future species distributions \cite{Wiersma2011}.
Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010}.
Additionally, textbooks introducing basic concepts of SDM in \emph{R} exists \cite{Guisan2017,Fletcher2019}.

@@

Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}.
Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}.
While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
Given its broad concept, many different measures of connectivity exist \cite{Kindlmann2008}.
On patch level, structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g., amount of suitable habitat) \cite{Kindlmann2008,With2019}.
Such measures are provided within the \emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
Furthermore, the \emph{lconnect} package \cite{Mestre2019} and \emph{Makurhini} package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
Another way to describe connectivity is based on graph theory with the advantage that functional connectivity can also be included \cite{Kindlmann2008}.
In graph theory \cite{Laita2011}, landscapes are described by nodes (i.e., habitat patches) connected by and functional connections called links (or edges) \cite{Laita2011}.
The \emph{grainscape} package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
More general, the \emph{igraph} package \cite{Csardi2006} provides functionality related to graph theory.
Resistance surfaces and least-cost paths are other tools to model functional connectivity.
The resistance surface describes the effects of facilitating or hindering the landscape's characteristics for an organism moving within it \cite{Adriaensen2003}.
Least-cost paths can be calculated using the \emph{gdistance} package \cite{vanEtten2017}.
Absorbing Markov chains that combine random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published \emph{samc} package \cite{Marx2020}.

@@

Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift, and selection \cite{Manel2003}.
Such insights improve our understanding of metapopulation dynamics, speciation, species' distributions, and conservation \cite{Storfer2007}.
By explicitly including landscape characteristics, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}.
As a result of its interdisciplinary, landscape genetics combines methods from different disciplines, including landscape ecology, spatial statistics, geography, and population genetics \cite{Storfer2007}.

Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also important for landscape genetics.
Further functionality for landscape genetics can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.

>>>>

\institute{
        Maximillian H.K. Hesselbarth \at
     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA \\
     \email{\href{mailto:mhk.hesselbarth@gmail.com}{\nolinkurl{mhk.hesselbarth@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Jakub Nowosad \at
     Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland \\
     \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Johannes M. Signer \at
     Department of Wildlife Sciences, University of Göttingen, Büsgenweg 3, 37077 Göttingen, Germany \\
     \email{\href{mailto:j.m.signer@gmail.com}{\nolinkurl{j.m.signer@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Laura J. Graham \at
     Geography, Earth \& Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom \\
     \email{\href{mailto:l.graham@bham.ac.uk}{\nolinkurl{l.graham@bham.ac.uk}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows use of the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
This democratizes scientific research as free-of-cost software removes one gatekeeper for researchers without access to proprietary software.

@@

The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review, we focus on software implemented in the \emph{R} programming language.
For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
Because in addition to handling spatial data, tasks such as statistical modeling, creation of publication-ready figures, and even preparation of complete reports can be done in the same working environment (Fig. \ref{fig:fig-workflow}), the \emph{R} programming language can be a very powerful tool.
Thus, in the first part, we present an overview of existing \emph{R} packages for landscape ecology (Table \ref{tab:packages}).
In the second part, we present a survey in which we asked the community how they currently use the \emph{R} programming language and to identify topics for which \emph{R} packages are presently missing for landscape ecology.

@@

In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
Currently, the most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
A \emph{raster} successor, \emph{terra}, aimed at the simpler interface and improved performance is now being developed \cite{R-terra}; however, it could take several years for this package to be adopted by other developers and users.
Alternatively, the \emph{stars} package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
Additionally, there are packages that improve some basic raster operations in terms of computational performance or compatibility between raster and vector operations, such as \emph{fasterize} \cite{Ross2020}, \emph{rasterDT} \cite{OBrien2020}, or \emph{exactextractr} \cite{Baston2020}.

@@

Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
A lot of this data is publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \emph{R} session.
Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows users to download data from the Landsat 8, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
Also, the \texttt{getData()} function from the \emph{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package \emph{geodata} ).
Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps .
This is also referred to as the spatial projection and is often the first barrier in spatial data analysis.
It is not only required to have all of the used data in the same projection, but also to select a proper CRS.
This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion, and different CRS are optimized for different properties, regions of the world, and scales .
Coordinates in spatial data represent one of many coordinate reference systems.
Two main groups of CRS, namely geographical and projected, exist, with each having many members.
In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
However, most landscape ecology studies should utilize projected CRSs, which use some measurement units (e.g., meters).
The selection of projected CRS should be based on the property of spatial data that we want to keep intact (e.g., does not distort areas, shapes, distances, or angles) and be appropriate for a given study area.
Tools to find an appropriate CRS for a certain region can be found at \url{https://spatialreference.org}, \url{https://epsg.org}, or \url{http://epsg.io}.
All packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations, allowing unification of spatial projections when the used data have different CRS.

A second common spatial data processing task is required when the available data extends over a larger area than the study region.
In this case, the pre-processing of spatial data should include vector clipping or raster cropping.
Related to that, masking certain areas of the study region using spatial filters (e.g.~water bodies, urban areas) can be required.
Packages from Section \ref{sec:spatrep} also allow for these operations.
Additionally, they offer many other operations, such as merging or joining spatial data, extracting values from one dataset into another, raster resolution changes, or vector data simplifications.
A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the \emph{grainchanger} package \cite{Graham2019}.
Furthermore, \emph{landscapetools} is a collection of various utility functions for the raster data model \cite{Sciaini2018}.

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine. Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.

@@

The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.

Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{R-mapview}.
Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.

@@

One of the most fundamental steps of landscape ecology analyses is to describe and quantify landscape characteristics \cite{Turner2005,Lausch2015}.
For discrete land cover classes, the composition (number and abundance) and configuration (spatial arrangement) of the landscape are often described using landscape metrics \cite{Gustafson1998,Uuemaa2009,Uuemaa2013,Gustafson2019}.
These metrics allow the comparison of different landscapes, quantification of temporal and spatial landscape changes and investigation of interactions between landscape characteristics and ecological processes \cite{Uuemaa2009}.

The introduction of the \emph{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019}.
However, while \emph{FRAGSTATS} is free to download at the developers' homepage, it is not open-source software.
This restricts its transparency, reproducibility, and extensibility.
Recently the \emph{landscapemetrics} package \cite{Hesselbarth2019a} was developed to provide a truly open-source solution to calculate landscape metrics.
The package allows calculation of the most widely used landscape metrics in a transparent and reproducible workflow within the \emph{R} environment.

More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
The \emph{geodiv} package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
Additionally, the \emph{belg} package allows calculation of the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.

Most landscape metrics are a represented by a single number depicting specific characteristics of a local landscape.
Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g.~Euclidian, Manhattan, Mahlanobis).
This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
All of the methods mentioned above are implemented in the \emph{motif} package \cite{R-motif}.

@@

Species distribution modeling (SDM) examines how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species' distributions, mainly to infer ecological processes and predict future species' distributions \cite{Wiersma2011}.
Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010}.
Additionally, textbooks introducing basic concepts of SDM in \emph{R} exists \cite{Guisan2017,Fletcher2019}.

@@

Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}.
Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}.
While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species such as habitat associations and dispersal distances \cite{Tischendorf2000,With2019}.
Given its broad concept, many different measures of connectivity exist \cite{Kindlmann2008}.
At the patch level, structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g., amount of suitable habitat) \cite{Kindlmann2008,With2019}.
Such measures are provided within the \emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
Furthermore, the \emph{lconnect} package \cite{Mestre2019} and \emph{Makurhini} package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
Another way to describe connectivity is based on graph theory with the advantage that functional connectivity can also be included \cite{Kindlmann2008}.
In graph theory \cite{Laita2011}, landscapes are described by nodes (i.e., habitat patches) connected by and functional connections called links (or edges) \cite{Laita2011}.
The \emph{grainscape} package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
More generally, the \emph{igraph} package \cite{Csardi2006} provides functionality for graph theoretic analyses.
Resistance surfaces and least-cost paths are other tools to model functional connectivity which include attributes of the matrix.
The resistance surface describes the effects of facilitating or hindering the landscape's characteristics for an organism moving within it \cite{Adriaensen2003}.
Least-cost paths can be calculated using the \emph{gdistance} package \cite{vanEtten2017}.
Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019a}, and is provided by the recently published \emph{samc} package \cite{Marx2020}.

@@

Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift, and selection \cite{Manel2003}.
Such insights improve our understanding of metapopulation dynamics, speciation, species' distributions, and conservation \cite{Storfer2007}.
By explicitly including landscape characteristics, landscape genetics provides a more detailed analysis than more abstract concepts (e.g., metapopulation genetics) \cite{Holderegger2006}.
As a result of its interdisciplinarity, landscape genetics draws together methods from multiple fields, including landscape ecology, spatial statistics, geography, and population genetics \cite{Storfer2007}.

Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also important for landscape genetics.
Further functionality for landscape genetics such as \ldots{} can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.
